Stop Lying to Me: New Visual Tools to Choose the Most Honest Nonlinear Dimension Reduction

### 2025-08-09 paper.qmd
Updated a1, a2 equality signs
Fixed Gammars and typos suggested by reviewers

Many new methods for NLDR have emerged in recent years, all designed to better capture specific structures potentially existing in \pD{}. Here we focus on five currently popular techniques: tSNE, UMAP, PHATE, TriMAP and PaCMAP. The methods tNSE, UMAP, TriMAP and PaCMAP can be considered to produce the \kD{} representation by minimizing the divergence between two inter-point distance distributions. PHATE is an example of a diffusion process spreading to capture geometric shapes, that include both global and local structure. (See @coifman2005 for an explanation of diffusion processes.)

@@

hex_grid_two_curvy <- ggplot(
  data = hex_grid_with_counts, 
  aes(x = x, y = y)) +
  geom_polygon(colour = "grey70", 
               aes(group = hex_poly_id), 
               fill = "#ffffff") +
  geom_point(data = tsne_two_curvy_scaled, 
             aes(x = emb1, y = emb2), 
             alpha = 0.5, size = 0.5, colour = clr_choice) +
  xlim(sc_xlims) + ylim(sc_ylims) +
  interior_annotation("b", sc_ltr_pos, cex = 2)

hex_grid_coloured_two_curvy1 <- ggplot() + 
  geom_polygon(
    data = hex_grid_with_counts, 
    aes(x = x, y = y,
      group = hex_poly_id, 
      fill = w_h), colour = "grey70", linewidth=0.2) +
  geom_point(data = tsne_two_curvy_scaled,
             aes(x = emb1, y = emb2),
             colour = clr_choice,
             alpha = 0.3,
             size = 0.5) +
  scale_fill_viridis_c(direction = -1, 
    na.value = "#ffffff", option = "E") +
  interior_annotation("b", sc_ltr_pos, cex = 2) + 
  xlim(sc_xlims) + ylim(sc_ylims) 
```

@@

hex_grid_nonempty_two_curvy <- ggplot(
  data = hex_grid_nonempty, 
  aes(x = x, y = y)) +
  geom_polygon(colour = "grey70", 
               aes(group = hex_poly_id), 
               fill = "#ffffff") +
  geom_point(data = df_bin_centroids_two_curvy, 
             aes(x = c_x, y = c_y), 
             colour = "#000000",
             size = 1) +
  xlim(sc_xlims) + ylim(sc_ylims) +
  interior_annotation("c", sc_ltr_pos, cex = 2) 
```

@@

To illustrate the method, we use \sD{} simulated data, which we call the "2NC7" data. It is has two separated nonlinear clusters, one forming a \gD{} curved shape, and the other a \tD{} curved shape, each consisting of $1000$ observations. The first four variables hold this cluster structure, and the remaining three are purely noise. We would consider $T=(X_1, X_2, X_3, X_4)$ to be the geometric structure (true model) that we hope to capture.

@@

$$
a_2 > \frac{r_2 + q(1 + r_2)}{(b_2 - 1)}.
$$

@@

$$
a_1 > \frac{2[r_2 + q(1 + r_2)]}{\sqrt{3}(b_2 - 1)}.
$$

@@

                            group = hex_poly_id),

@@

hex_grid_coloured_two_curvy2 <- ggplot() + 
  geom_polygon(
    data = hex_grid_with_counts_two_curvy2, 
    aes(x = x, y = y, 
        group = hex_poly_id, 
        fill = w_h), 
        colour = "grey70", 
        linewidth=0.2) +
  geom_point(data = tsne_two_curvy_scaled,
           aes(x = emb1, y = emb2),
           alpha = 0.3,
           size = 0.5) +
  scale_fill_viridis_c(direction = -1, 
    na.value = "#ffffff", option = "C") +
  xlim(sc_xlims) + ylim(sc_ylims) +
  interior_annotation("a", sc_ltr_pos, cex = 2)

@@

hex_grid_coloured_two_curvy3 <-  ggplot() + 
  geom_polygon(
    data = hex_grid_with_counts_two_curvy3, 
    aes(x = x, y = y, 
        group = hex_poly_id, 
        fill = w_h), 
        colour = "grey70", linewidth=0.2) +
  geom_point(data = tsne_two_curvy_scaled,
           aes(x = emb1, y = emb2),
           alpha = 0.3,
           size = 0.5) +
  scale_fill_viridis_c(direction = -1, na.value = "#ffffff", option = "D") +
  xlim(sc_xlims) + ylim(sc_ylims) +
  interior_annotation("c", sc_ltr_pos, cex = 2)
```

```{r}
#| label: best-fit-tsne
#| fig-cap: "Lifting the \\gD{} fitted model into \\pD{}. Two projections of the \\pD{} fitted model overlaying the data are shown in b, c. The fit is reasonably tight with the data in one cluster (top one in b), but slightly less so in the other cluster probably because it is $3\\text{-}D$. Notice also that, in the \\gD{} layout the two clusters have internal gaps which creates a model with some holes. This lacy pattern happens regardless of the hyper-parameter choice, but this doen't severely impact the \\pD{} model representation."
#| fig-width: 8
#| fig-height: 2

@@

Comparing the RMSE to obtain the best fit is appropraite if the same NLDR method is used. However, because the RMSE is computed on \pD{} data it measures the fit between model and data so it can also be used to compare the fit of different NLDR methods. A lower RMSE indicates a better NLDR representation.

>>>>

Many new methods for NLDR have emerged in recent years, all designed to better capture specific structures potentially existing in \pD{}. Here we focus on five currently popular techniques: tSNE, UMAP, PHATE, TriMAP and PaCMAP. The methods tNSE, UMAP, TriMAP and PaCMAP can be considered for producing the \kD{} representation by minimizing the divergence between two inter-point distance distributions. PHATE is an example of a diffusion process spreading to capture geometric shapes, that include both global and local structure. (See @coifman2005 for an explanation of diffusion processes.)

@@

hex_grid_two_curvy <- ggplot(
  data = hex_grid_with_counts, 
  aes(x = x, y = y)) +
  geom_polygon(colour = "grey70", 
               aes(group = h), 
               fill = "#ffffff") +
  geom_point(data = tsne_two_curvy_scaled, 
             aes(x = emb1, y = emb2), 
             alpha = 0.5, size = 0.5, colour = clr_choice) +
  xlim(sc_xlims) + ylim(sc_ylims) +
  interior_annotation("b", sc_ltr_pos, cex = 2)

hex_grid_coloured_two_curvy1 <- ggplot() + 
  geom_polygon(
    data = hex_grid_with_counts, 
    aes(x = x, y = y,
      group = h, 
      fill = w_h), colour = "grey70", linewidth=0.2) +
  geom_point(data = tsne_two_curvy_scaled,
             aes(x = emb1, y = emb2),
             colour = clr_choice,
             alpha = 0.3,
             size = 0.5) +
  scale_fill_viridis_c(direction = -1, 
    na.value = "#ffffff", option = "E") +
  interior_annotation("b", sc_ltr_pos, cex = 2) + 
  xlim(sc_xlims) + ylim(sc_ylims) 
```

@@

hex_grid_nonempty_two_curvy <- ggplot(
  data = hex_grid_nonempty, 
  aes(x = x, y = y)) +
  geom_polygon(colour = "grey70", 
               aes(group = h), 
               fill = "#ffffff") +
  geom_point(data = df_bin_centroids_two_curvy, 
             aes(x = c_x, y = c_y), 
             colour = "#000000",
             size = 1) +
  xlim(sc_xlims) + ylim(sc_ylims) +
  interior_annotation("c", sc_ltr_pos, cex = 2) 
```

@@

To illustrate the method, we use \sD{} simulated data, which we call the "2NC7" data. It has two separated nonlinear clusters, one forming a \gD{} curved shape, and the other a \tD{} curved shape, each consisting of $1000$ observations. The first four variables hold this cluster structure, and the remaining three are purely noise. We would consider $T=(X_1, X_2, X_3, X_4)$ to be the geometric structure (true model) that we hope to capture.

@@

$$
a_2 \geq \frac{r_2 + q(1 + r_2)}{(b_2 - 1)}.
$$

@@

$$
a_1 \geq \frac{2[r_2 + q(1 + r_2)]}{\sqrt{3}(b_2 - 1)}.
$$

@@

                            group = h),

@@

hex_grid_coloured_two_curvy2 <- ggplot() + 
  geom_polygon(
    data = hex_grid_with_counts_two_curvy2, 
    aes(x = x, y = y, 
        group = h, 
        fill = w_h), 
        colour = "grey70", 
        linewidth=0.2) +
  geom_point(data = tsne_two_curvy_scaled,
           aes(x = emb1, y = emb2),
           alpha = 0.3,
           size = 0.5) +
  scale_fill_viridis_c(direction = -1, 
    na.value = "#ffffff", option = "C") +
  xlim(sc_xlims) + ylim(sc_ylims) +
  interior_annotation("a", sc_ltr_pos, cex = 2)

@@

hex_grid_coloured_two_curvy3 <-  ggplot() + 
  geom_polygon(
    data = hex_grid_with_counts_two_curvy3, 
    aes(x = x, y = y, 
        group = h, 
        fill = w_h), 
        colour = "grey70", linewidth=0.2) +
  geom_point(data = tsne_two_curvy_scaled,
           aes(x = emb1, y = emb2),
           alpha = 0.3,
           size = 0.5) +
  scale_fill_viridis_c(direction = -1, na.value = "#ffffff", option = "D") +
  xlim(sc_xlims) + ylim(sc_ylims) +
  interior_annotation("c", sc_ltr_pos, cex = 2)
```

```{r}
#| label: best-fit-tsne
#| fig-cap: "Lifting the \\gD{} fitted model into \\pD{}. Two projections of the \\pD{} fitted model overlaying the data are shown in b, c. The fit is reasonably tight with the data in one cluster (top one in b), but slightly less so in the other cluster probably because it is $3\\text{-}D$. Notice also that, in the \\gD{} layout the two clusters have internal gaps which creates a model with some holes. This lacy pattern happens regardless of the hyper-parameter choice, but this doesn't severely impact the \\pD{} model representation."
#| fig-width: 8
#| fig-height: 2

@@

Comparing the RMSE to obtain the best fit is appropriate if the same NLDR method is used. However, because the RMSE is computed on \pD{} data it measures the fit between model and data so it can also be used to compare the fit of different NLDR methods. A lower RMSE indicates a better NLDR representation.
